---
title: "Chip_p300_mESC"
author: "Viviana"
date: '2022-03-23'
output: html_document
editor_options: 
  chunk_output_type: console
---

# Assignment 3 and 4 

## Week 3 

1. Download a mouse ChIPseq dataset (see suggestion below)
2. Process it from the raw data, obtaining:
bam file, along with number and percentage of mapped reads
bigwig file
peaks
3. How many peaks do you find?
4. Plot the signal round one of the peaks


```{r}
suppressPackageStartupMessages({
  library(AnnotationHub)
  library(Rsubread)
  library(rtracklayer)
  library(Biostrings)
  library(Rfastp)
  library(epiwraps)
})
ah <- AnnotationHub()


```


```{r, eval=FALSE}

dir.create("raw")
options(timeout=1800)
download.file("https://www.encodeproject.org/files/ENCFF001LJN/@@download/ENCFF001LJN.fastq.gz" , dest="raw/p300.fastq.gz")

```



```{r, eval=FALSE}
dir.create("rfastp.trimmed")
qc <- Rfastp::rfastp("raw/p300.fastq.gz", outputFastq="rfastp.trimmed/p300", 
                     maxNfilter=0, thread=4, overrepresentationAnalysis=TRUE)
```



```{r, eval=FALSE}
 
genome <- ah[["AH49772"]]
#AH53222 ensbd 
#AH49772 Two bits 
dir.create("GRCm38_genome")

export(import.2bit(genome), "GRCm38_genome/genome.fasta.gz", compress=TRUE)

Rsubread::buildindex("GRCm38_genome/rsubread", reference="GRCm38_genome/genome.fasta.gz")
```



```{r, eval=FALSE}
dir.create("aligned")
align.stats <- Rsubread::align(index="GRCm38_genome/rsubread", type="dna",
                               output_file="aligned/p300.bam",
                               readfile1="rfastp.trimmed/p300_R1.fastq.gz", 
                               nthreads=4, sortReadsByCoordinates=TRUE)
align.stats
```




```{r}

ensdb <- ah[["AH89211"]]
plotSignalTracks(list("aligned/p300.bam"), region="Atf3", ensdb=ensdb, tyranscripts= "full")
```


```{r, eval=FALSE}
#BiocManager::install("MACSr")
library(MACSr)
dir.create("peaks")

download.file("https://www.encodeproject.org/files/ENCFF460EOA/@@download/ENCFF460EOA.bed.gz", "peaks/p300.bed.gz")


```


```{r, eval=FALSE}
dir.create("tracks2")
#bam2bw("aligned/p300.bam", "tracks/p300_full.bw", binWidth=1, extend=50, scaling=TRUE)
#bam2bw("aligned/p300.bam", "tracks/p300.bw", binWidth=10, extend=50, scaling=TRUE)
download.file("https://www.encodeproject.org/files/ENCFF552AJY/@@download/ENCFF552AJY.bigWig", "tracks2/p300.bw")

```



```{r}
p300 <- import("peaks/p300.bed.gz", format="NarrowPeak")
plotSignalTracks(list(signal="tracks2/p300.bw", peaks=p300),region=p300[1], ensdb=ensdb)
```



##Week 4

Complete last week’s assignment
-Find and download the peak-level data (i.e. bed-like format) for the following histone modifications in mouse embryonic stem cells (mESC) from ENCODE:

H3K4me3, H3K4me1, H3K27ac, and H3K27me3

-Of the p300 peaks you identified last week, what proportion overlap each of the marks?
Please make sure that you name your final file assignment.html !


```{r , eval=FALSE}

# downlowad peaks and tracks 
# H3K27ac
download.file("https://www.encodeproject.org/files/ENCFF274UIB/@@download/ENCFF274UIB.bed.gz", "peaks/H3K27ac.bed.gz")
download.file("https://www.encodeproject.org/files/ENCFF121THA/@@download/ENCFF121THA.bigWig", "tracks/H3K27ac.bw")


# H3K27me3
download.file("https://www.encodeproject.org/files/ENCFF008XKX/@@download/ENCFF008XKX.bed.gz", "peaks/H3K27me3.bed.gz")
download.file("https://www.encodeproject.org/files/ENCFF160FEV/@@download/ENCFF160FEV.bigWig", "tracks/H3K27me3.bw")


# H3K4me1
download.file("https://www.encodeproject.org/files/ENCFF333IJH/@@download/ENCFF333IJH.bed.gz", "peaks/H3K4me1.bed.gz")
download.file("https://www.encodeproject.org/files/ENCFF016YZA/@@download/ENCFF016YZA.bigWig", "tracks/H3K4me1.bw")


# H3K4me3
download.file("https://www.encodeproject.org/files/ENCFF247GVM/@@download/ENCFF247GVM.bed.gz", "peaks/H3K4me3.bed.gz")
download.file("https://www.encodeproject.org/files/ENCFF611GSQ/@@download/ENCFF611GSQ.bigWig", "tracks/H3K4me3.bw")




```


```{r}
k27ac <- import("peaks/H3K27ac.bed.gz", format="NarrowPeak")

k27me3 <- import("peaks/H3K27me3.bed.gz", format="NarrowPeak")

k4me1 <- import("peaks/H3K4me1.bed.gz", format="NarrowPeak")

k4me3 <- import("peaks/H3K4me3.bed.gz", format="NarrowPeak")

```





```{r}
#to download the peaks 




head(p300)

#to change format of seq name 
# number of  peaks 
length(p300) #28341
length(k27ac)#49716
length(k27me3)#12621
length(k4me1)#85881
length(k4me3)#52499

# how many oeaks overlap 
#H3K27ac 
peakOverlap_k27ac<- overlapsAny(p300, k27ac)
table(peakOverlap_k27ac)#14699 

#H3K27me3 
peakOverlap_k27me3<- overlapsAny(p300, k27me3)
table(peakOverlap_k27me3)#82

#H3K4me1
peakOverlap_k4me1<- overlapsAny(p300, k4me1)
table(peakOverlap_k4me1)#7919


#H3K4me3
peakOverlap_k4me3<- overlapsAny(p300, k4me3)
table(peakOverlap_k4me3)#10341





```


Comment: Finally I was able to solve the problem I was having with the peak download, now everything seemed to work well. 