---
title: "assignment 6"
author: "Viviana"
date: '2022-04-06'
output: html_document
editor_options: 
  chunk_output_type: console
---


# Assignment 6_ update 

Previously I misunderstood question number b, therefore I rewrote and I am correcting the file! 

1. Choose a transcription factor, e.g. CREB1, REST, GATA5, EGR1, GCR (or any of your choice that has a motif and available ChIPseq data)
2. Download the (e.g. Mouse) peaks for that factor (whatever cell type)
3. Identify the instances of the factor’s motif

Answer the following questions:
a. Of all the peaks, what proportion contains a motif for the factor? (Expected form of an answer: "of the XX peaks, XX (XX%) contain a motif")
b. Of all instances of that motif in the genome, what proportion is bound by the factor (i.e. has a peak)? (Expected form of an answer: "of the XX motif instances, XX (XX%) overlap a peak")

Don't forget to render your markdown and push it as assignment.html !

```{r}
suppressPackageStartupMessages({
  library(GenomicRanges)
  library(memes)
  library(Biostrings)
  library(AnnotationHub)
  library(MotifDb)
  library(universalmotif)
  library(PWMEnrich)
  library(ggplot2)
})
```

I will use Gata1 on mm10 data on G1E cell line  

```{r}
download.file("https://www.encodeproject.org/files/ENCFF978KGB/@@download/ENCFF978KGB.bed.gz", "creb_ENCFF978KGB.bed.gz")
peaks <- rtracklayer::import("creb_ENCFF978KGB.bed.gz", format="NarrowPeak")
seqlevelsStyle(peaks) <- "Ensembl"
peaks_chr1 <- peaks[seqnames(peaks)=="1"]

```




```{r}

peak_centers <- resize(peaks_chr1, fix="center", width=100)
# we get the genome sequence:
ah <- AnnotationHub(localHub=FALSE)

genome <- ah[["AH49723"]]
# we get the sequences corresponding to the peak centers:
peak_seqs <- memes::get_sequence(peak_centers, genome)
length(peak_centers)
length(peak_seqs)
peak_seqs

```





```{r}

#gadem <- rGADEM::GADEM(peak_seqs, genome=genome)

# we search for "Gata1" in the motif database
motifs <- query(MotifDb, "CREB1")
head(motifs)
# there are several matching motifs:
names(motifs)
# we select one:
motif <- motifs[["Hsapiens-jaspar2016-CREB1-MA0018.1"]]
# we visualize it:
view_motifs(motif)


```




```{r}
#' findMotifInstances
#'
#' A wrapper around TFBSTools for scanning motif occurence, handling the 
#' coordinate conversion as `memes` does.
#' 
#' @param seqs A set of sequences, e.g. `DNAStringSet`, optionally with 
#'   coordinate as names to enable conversion.
#' @param motif A motif, in any format recognized by `universalmotif`
#' @param keepMatchedSeq Logical; whether to keep the matched sequence.
#' @param ... Passed to `TFBSTools::searchSeq`; can for instance be used to set
#'   the number of threads to use, e.g. with `mc.cores=2`
#'
#' @return A `GRanges` object
findMotifInstances <- function(seqs, motif, keepMatchedSeq=FALSE, ...){
  motif <- universalmotif::convert_motifs(motif, "TFBSTools-PWMatrix")
  x <- suppressWarnings(TFBSTools::searchSeq(motif, subject=seqs))
  peaks <- strsplit(gsub("-",":",names(seqs)), ":")
  if(all(lengths(peaks)==3)){ # convert relative coordinates to absolute
    chrs <- sapply(peaks,FUN=function(x) x[1])
    offsets <- sapply(peaks,FUN=function(x) as.integer(x[2]))-1L
    i <- rep(seq_along(x),lengths(x))
    return(GRanges(chrs[i], 
                   IRanges(offsets[i]+as.integer(unlist(lapply(x, start))),
                           offsets[i]+as.integer(unlist(lapply(x, end)))),
                   strand=unlist(lapply(x, strand)),
                   score=as.numeric(unlist(lapply(x, FUN=function(x) score(x))))
                   ))
  }
  x <- as(x, "GRanges")
  score(x) <- x$absScore
  keepFields <- c("score","relScore")
  if(keepMatchedSeq) keepFields <- c(keepFields, "siteSeqs")
  mcols(x) <- mcols(x)[,intersect(colnames(mcols(x)), keepFields)]
  x
}
```



```{r}
moi2 <- findMotifInstances(peak_seqs, motif, mc.cores=2)
head(moi2)

length(peak_seqs)#2380
length(moi2)#2015
```




```{r}

overlaps<-overlapsAny(peak_centers, moi2)
a<-table(overlaps) # true overlaps 1364 

length(peak_centers)#2015

prop.table(a)

#true 0.6769231 


overlaps2<-overlapsAny( moi2, peak_centers)
b<-table(overlaps2) # true overlaps 1364 
length(moi2)#2380

mouseseq <- import(genome, "2bit", which= as(seqinfo(genome), "GenomicRanges"))
motif_genome <- findMotifInstances(mouseseq, motif, mc.cores=2)

length(motif_genome)#11817058 

overlap_c<-overlapsAny(motif_genome, peak_centers)

c<-table(overlap_c)
prop.table(c) #TRUE 0.0002350839 0.0235%





```



Answer to the questions:

a. Of the 2015 peaks of CREB1 in chromosome 1, 1364 (67.7%) contains the selected motif for the factor 

b. The selected motif for CREB1 has 2380 instances in chromosome 1, of those 2380 (100%) overlap with a peak for CREB1 binding. The peak centers are lower than the motif instances, this is possible since it could be that there are multiple motif instances in the sampe peak_seq. And they are all bound to the factor. From this data it seems that there are no motif instances that are not bound by CREB1 in chromosome 1 

c. of all the instances of the CREB1 motif in the genome 11817058 , 2778 overlap a peak (0.0235%) of CREB1 at chromosome1 

